Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-04-16 paper/GEB Submission/manuscript_GEB.tex
more edits, hooray

\begin{abstract}
Aim: Despite unprecedented environmental change due to anthropogenic
pressure, recent work has found increasing dissimilarity due to turnover
but no overall trend in species diversity through time at the local
scale. Functional diversity provides a potentially powerful alternative
approach for understanding community composition by linking shifts in
species identity to mechanisms of ecosystem processes. Here we present
the first multi-taxa, multi-system analysis of functional change through
time.

@@

Methods: We paired thousands of bird and mammal assemblage time series
from the BioTIME database with existing trait data representative of a
species' functional role to reconstruct time series of functional
diversity metrics. Using generalized linear mixed models, we estimated
general trends in those metrics and trends for individual studies.

Results: We found no overall trend in any functional diversity metric,
which held even after correcting for changes in species richness. At the
study-level, there was substantial heterogeneity in the direction and
type of functional change exhibited. None of the included study
characteristics such as taxa, realm, biome, or protection status were
significantly related to the type of change exhibited by a community.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. At the study-level, the majority group of
studies showed no species or functional trends, but other types of
change were found for multiple studies, including studies experiencing
changes in functional redundancy, increased species and functional
richness, or loss of functional richness. With no one prevailing
scenario of change, it will be critical to link change scenarios to
ecological context.
\end{abstract}

@@

We have stronger expectations for changes in the prevalence of some
individual traits. For example, animal body size is expected to decrease
as a result of climate change, a phenomena that has been documented in
multiple taxa empirically and experimentally (37--41). There is some
evidence that this holds true for birds (42), but the picture is likely
more complicated for mammals, where urbanization may actually lead to
larger body sizes due to novel food sources (43), even as megafaunal
loss leads to on average smaller community body size. For dietary
traits, recent work documenting insect declines (citation) points to
potentially significant negative impacts on insectivorous animals
(44--46). Predicted extinctions based on species-level vulnerability
point to further dietary shifts, favoring increases in invertivorous
species (47). Some systems also show significant shifts in the
prevalence of different kinds of foragers in birds, for example loss of
arboreal foragers in agricultural systems (48), and loss of neotropical
understory foragers even in protect areas (49).

@@

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
We further evaluate evidence for key predictions of individual trait
prevalence from the literature, including changes in mean body size,
dietary traits, and foraging strata.`

@@

We also calculated the community-weighted mean (\emph{CWM}) of all
continuous traits to examine \textcolor{blue}{turnover} in the
distribution of each trait.
\textcolor{blue}{Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained. CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

@@

We found no significant overall \textcolor{blue}{temporal} trend in
species richness or functional diversity metrics
\textcolor{blue}{including functional richness, evenness, or divergence}
(observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall \textcolor{blue}{temporal} trends
indicate that although some studies experience increasing or decreasing
trends, the average trend across studies
\textcolor{blue}{was not significantly different from zero}` (Table
\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
different taxa, biomes, realms, or protection statuses were also
non-significant for richness and functional diversity metrics, with the
exception of a significantly increasing trend for functional evenness of
marine studies, and a significantly decreasing functional richness slope
for mammal studies. However, trends were not exhibited for the
\textcolor{blue}{corrected} metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. The general trends for \emph{CWM} models
were similarly not significant, with a significant positive trend for
only percentage of fish in diet composition.

@@

\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
(66--68).
\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
(66).
\textcolor{blue}{In general, functional capacity of mammals showed greater declines that bird communities.}

@@

\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or simply unexpected. For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation. The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas. Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower }(52).
\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts. Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}

\textcolor{blue}{We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
(69),
\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(57,
70).
\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(71).
\textcolor{blue}{An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world. Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(72).

Our results were also inconsistent with all predictions for changes in
trait prevalence, including foraging strata, body size, and dietary
traits.
\textcolor{blue}{We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe. We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data. Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
(73).
\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
(74).

Rather than declines in insectivory and increases in invertivory
predicted by previous work, we instead found changes in the degree of
consumption of fish, nectar, and vertebrates.
\textcolor{blue}{Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, which makes these results suprising. Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}

@@

\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens. The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level. In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend. While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover. We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.}

@@

\textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy. By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella. For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(76).
\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(77).
\textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
(78).

\textcolor{blue}{The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover. While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space. These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function. Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden). Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
(35)
\textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
(79--81).

@@

The addition of functional trait data to the biodiversity change
conversation is yet another illustration of how critical context is for
understanding ecological patterns (86, 87).
\textcolor{blue}{Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer. Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness. The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, the main barrier for taking a functional approach to biodiversity change.}

@@

\leavevmode\vadjust pre{\hypertarget{ref-catford2022}{}}%
\CSLLeftMargin{86. }%
\CSLRightInline{Catford JA, Wilson JRU, Pyšek P, Hulme PE, Duncan RP
(2022) \href{https://doi.org/10.1016/j.tree.2021.09.007}{Addressing
context dependence in ecology}. \emph{Trends in Ecology \& Evolution}
37(2):158--170.}

\leavevmode\vadjust pre{\hypertarget{ref-spake2023}{}}%
\CSLLeftMargin{87. }%
\CSLRightInline{Spake R, et al. (2023)
\href{https://doi.org/10.1111/brv.12939}{Understanding {`}it depends{'}
in ecology: a guide to hypothesising, visualising and interpreting
statistical interactions}. \emph{Biological Reviews} 98(4):983--1002.}

\leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
\CSLLeftMargin{88. }%
\CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
M-AR (2020) North american breeding bird survey dataset 1966 - 2019,
version 2019.0.
doi:\href{https://doi.org/10.5066/P9J6QUF6}{10.5066/P9J6QUF6}.}

\leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
\CSLLeftMargin{89. }%
\CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
SEAMAP Available at: \url{http://www.iobis.org/mapper/?dataset=2245}.}

\leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
\CSLLeftMargin{90. }%
\CSLRightInline{Silva FR da Brazil dataset 1.}

\leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
\CSLLeftMargin{91. }%
\CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
study. Available at: \url{http://srelherp.uga.edu/projects/rbay.htm}.}

\leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
\CSLLeftMargin{92. }%
\CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
amphibia at northwest region of sao paulo state, brazil: Microhabitat,
seasonality, diet and multidimensional niche. PhD thesis.}

\leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
\CSLLeftMargin{93. }%
\CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
oosterschelde before, during and after the storm-surge barrier
(1982{\textendash}1990). Available at:
\url{http://www.iobis.org/mapper/?dataset=505}.}

\leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
markakol depression (southern altai). (249):3--15.}

\leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
(2004) \href{https://doi.org/10.34080/os.v14.20236}{The 37-year dynamics
of a subalpine passerine bird community, with special emphasis on the
influence of environmental temperature and epirrita autumnata cycles}.
\emph{Ornis Svecica} 14(3):63--106.}

\leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{Gido K (2019) CFP01 Fish population on selected
watersheds at Konza Prairie.
doi:\href{https://doi.org/10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}{10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}.}

\leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{Hogstad O (1993)
\href{https://www.jstor.org/stable/23735355}{Structure and dynamics of a
passerine bird community in a spruce-dominated boreal forest. A 12-year
study}. \emph{Annales Zoologici Fennici} 30(1):43--54.}

\leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
\href{https://doi.org/10.2307/2937074}{Bird Community Dynamics in a
Temperate Deciduous Forest: Long-Term Trends at Hubbard Brook}.
\emph{Ecological Monographs} 56(3):201--220.}

\leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
marine mammal observation data, 1987{\textendash}2006. Available at:
\url{http://www.iobis.org}.}

\leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}

\leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
Migration of birds of prey in the mouth of the river irkut.
(108):3--17.}

\leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
Ministry of Environment of Japan (2013) Monitoring site 1000 shorebird
survey. Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html}.}

>>>>

\begin{abstract}
Aim: Despite unprecedented environmental change due to anthropogenic
pressure, recent work has found increasing dissimilarity due to turnover
but no overall trend in species diversity through time at the local
scale. Functional diversity provides a potentially powerful alternative
approach for understanding community composition by linking shifts in
species identity to the characteristics that confer ecosystem processes.
Here we present the first multi-taxa, multi-system analysis of
functional change through time.

@@

Methods: We paired thousands of bird and mammal assemblage time series
from the BioTIME database with existing trait data representative of a
species' functional role to reconstruct time series of functional
diversity and composition metrics. Using generalized linear mixed
models, we estimated general trends in those metrics and trends for
individual studies.

Results: We found evidence of decreasing functional richeness for mammal
communities and increasing functional evenness in marine communities,
but otherwise no overall trends in functional diversity metrics, which
held even after correcting for changes in species richness. With the
exception of those two trends, study characteristics such as taxa,
realm, biome, or protection status did not distinguish between types of
change exhibited by communities. At the study-level, there was
substantial heterogeneity in the direction and type of functional change
exhibited. We further did not find evidence to support multiple
predictions for inidividual traits, including decreasing body size,
dietary shifts, or changes in foraging strata.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevalant than the other across many
taxa, biomes, and realms. Loss of functional richness in mammal
communities indicates that they are on average faring worse than bird
communities. At the study-level, the majority group of studies showed no
species or functional trends, but other types of change were found for
multiple studies, including studies experiencing changes in functional
redundancy, increased species and functional richness, or loss of
functional richness. With no one prevailing scenario of change, it will
be critical to link change scenarios to ecological context.
\end{abstract}

@@

\textcolor{blue}{We have stronger expectations for changes in the prevalence of some individual traits. For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally}
(37--41).
\textcolor{blue}{There is some evidence that this holds true for birds }(42)\textcolor{blue}{, but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources}
(43)\textcolor{blue}{, even as megafaunal loss leads to on average smaller community body size. For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals}
(44--46).
\textcolor{blue}{Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species }(47).
\textcolor{blue}{Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems}
(48)\textcolor{blue}{, and loss of neotropical understory foragers even in protect areas}
(49).

@@

We assess thousands of mammal and bird functional diversity time series
to determine whether or not
\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities. We present a few areas of change consensus across communities, and even more scenarios where functional trends are unexplained and warrant further theoretical and experimental examination. We assess functional diversity trends at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies. We further evaluate evidence of key predictions for individual traits using trait-level compositional trends, including changes in mean body size, dietary traits, and foraging strata.}

@@

We also calculated the community-weighted mean (\emph{CWM}) of all
continuous traits to examine \textcolor{blue}{turnover} in the
distribution of each trait.
\textcolor{blue}{Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained. CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}

@@

We found no significant overall \textcolor{blue}{temporal} trend in
species richness or functional diversity metrics
\textcolor{blue}{including functional richness, evenness, or divergence}
(observed or \textcolor{blue}{corrected}) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall \textcolor{blue}{temporal} trends
indicate that although some studies experience increasing or decreasing
trends, the average trend across studies
\textcolor{blue}{was not significantly different from zero} (Table
\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
different taxa, biomes, realms, or protection statuses were also
non-significant for richness and functional diversity metrics, with the
exception of a significantly increasing trend for functional evenness of
marine studies, and a significantly decreasing functional richness slope
for mammal studies. However, trends were not exhibited for the
\textcolor{blue}{corrected} metric indicating that differences
\textcolor{blue}{in functional diversity metrics} were largely due to
changes in species richness. The general trends for \emph{CWM} models
were similarly not significant, with a significant positive trend for
only percentage of fish in diet composition.

@@

\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
(66--68).
\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
(66).
\textcolor{blue}{In general, functional capacity of mammals showed greater declines than bird communities.}

@@

\textcolor{blue}{The addition of functional trait data illuminated multiple areas where a synthesis approach is incongruous with system-specific studies or simply unexpected. For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation. The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas. Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower }(52).
\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts. Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}

\textcolor{blue}{We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
(69),
\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(57,
70).
\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(71).
\textcolor{blue}{An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world. Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work }(72).

\textcolor{blue}{Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits. We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe. We found no evidence of general trends in the CWM for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data. Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
(73).
\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
(74).

\textcolor{blue}{Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates. Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising. Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammals and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}

@@

\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens. The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level in both species and functional metrics. For example, it appears at the aggregate level that species turnover does not translate to functional turnover based on a lack of trends in CWM's. However at the study-level, all studies experiencing species turnover are also experiencing significant turnover for multiple individual traits, there is simply no consistency across communities in the kind of turnover occurring to be manifest in general trends.}

\textcolor{blue}{In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend. While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover. We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.}

@@

\textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy. By definition, if a community gains or loses species while functional diversity metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella. For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(76).
\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(77).
\textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
(78).

\textcolor{blue}{The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover. While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space. These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function. Notably, the studies that exhibited species richness increases were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden). Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
(35)
\textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
(79--81).

@@

\textcolor{blue}{The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023]. Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer. Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness. The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, the main barrier for taking a functional approach to biodiversity change.}

@@

\leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
\CSLLeftMargin{86. }%
\CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
M-AR (2020) North american breeding bird survey dataset 1966 - 2019,
version 2019.0.
doi:\href{https://doi.org/10.5066/P9J6QUF6}{10.5066/P9J6QUF6}.}

\leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
\CSLLeftMargin{87. }%
\CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
SEAMAP Available at: \url{http://www.iobis.org/mapper/?dataset=2245}.}

\leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
\CSLLeftMargin{88. }%
\CSLRightInline{Silva FR da Brazil dataset 1.}

\leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
\CSLLeftMargin{89. }%
\CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
study. Available at: \url{http://srelherp.uga.edu/projects/rbay.htm}.}

\leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
\CSLLeftMargin{90. }%
\CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
amphibia at northwest region of sao paulo state, brazil: Microhabitat,
seasonality, diet and multidimensional niche. PhD thesis.}

\leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
\CSLLeftMargin{91. }%
\CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
oosterschelde before, during and after the storm-surge barrier
(1982{\textendash}1990). Available at:
\url{http://www.iobis.org/mapper/?dataset=505}.}

\leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
\CSLLeftMargin{92. }%
\CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
markakol depression (southern altai). (249):3--15.}

\leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
\CSLLeftMargin{93. }%
\CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
(2004) \href{https://doi.org/10.34080/os.v14.20236}{The 37-year dynamics
of a subalpine passerine bird community, with special emphasis on the
influence of environmental temperature and epirrita autumnata cycles}.
\emph{Ornis Svecica} 14(3):63--106.}

\leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{Gido K (2019) CFP01 Fish population on selected
watersheds at Konza Prairie.
doi:\href{https://doi.org/10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}{10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}.}

\leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{Hogstad O (1993)
\href{https://www.jstor.org/stable/23735355}{Structure and dynamics of a
passerine bird community in a spruce-dominated boreal forest. A 12-year
study}. \emph{Annales Zoologici Fennici} 30(1):43--54.}

\leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
\href{https://doi.org/10.2307/2937074}{Bird Community Dynamics in a
Temperate Deciduous Forest: Long-Term Trends at Hubbard Brook}.
\emph{Ecological Monographs} 56(3):201--220.}

\leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
marine mammal observation data, 1987{\textendash}2006. Available at:
\url{http://www.iobis.org}.}

\leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}

\leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
Migration of birds of prey in the mouth of the river irkut.
(108):3--17.}

\leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
Ministry of Environment of Japan (2013) Monitoring site 1000 shorebird
survey. Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html}.}

\leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
Ministry of Environment of Japan (2014) Monitoring site 1000 village
survey - bird survey data (2005{\textendash}2012). Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html)}.}

\leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{Monitoring site 1000 village survey - medium and large
mammal survey data (2006{\textendash}2012) (2014) Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html}.}

\leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
\CSLLeftMargin{103. }%
\CSLRightInline{Preston FW (1960)
\href{https://doi.org/10.2307/1931793}{Time and space and the variation
of species}. \emph{Ecology} 41(4):612--627.}

\leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
\CSLLeftMargin{104. }%
\CSLRightInline{Svensson S (2006) Species composition and population
fluctuations of alpine bird communities during 38 years in the
scandinavian mountain range. \emph{Ornis Svecica} 16:183--210.}
